---
title: "Mini-Project"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(mosaic)
library(manipulate)
hate_crimes <- read.csv("/Users/jasoncheong/Downloads/Readings/CS/Mini-Project/hate_crimes.csv")
```

### Part 1: Macro Analysis

#### Hypothesis 1: Is a state with more Trump voters more likely to have more hate crimes?
```{r analysis, results = "hide"}
x <- select(hate_crimes, median_household_income:avg_hatecrimes_per_100k_fbi)
cor(x, use = "complete.obs") > 0.5 
cor(x, use = "complete.obs") < -0.5 
```
From the above, **only the Trump voters** variable is correlated with both the FBI and SPLC hate crimes data.

```{r}
cor(hate_crimes$share_voters_voted_trump, hate_crimes$avg_hatecrimes_per_100k_fbi, use = "complete.obs")

ggplot(data = hate_crimes, aes(x = share_voters_voted_trump, y = avg_hatecrimes_per_100k_fbi)) + geom_point() + labs(title = "Trump Voters in relation to Average Hate Crimes (2010-15)", x = "Share of Trump Voters", y = "Average Hate Crimes (2010-15)")
```
Correlation between Trump Voters and Hate Crimes (FBI): **-0.505**

```{r}
cor(hate_crimes$share_voters_voted_trump, hate_crimes$hate_crimes_per_100k_splc, use = "complete.obs")

ggplot(data = hate_crimes, aes(x = share_voters_voted_trump, y = hate_crimes_per_100k_splc)) + geom_point() + labs(title = "Trump Voters in relation to Hate Crimes (Election Week 2016)", x = "Share of Trump Voters", y = "Hate Crimes (Election Week 2016)")
```
Correlation between Trump Voters and Hate Crimes (FBI): **-0.657**

### Part 2: Micro Analysis

### Part 3: Case Studies

#### Case Study 1: Washington, D.C.
```{r}
hate_crimes %>% select(state, share_voters_voted_trump, avg_hatecrimes_per_100k_fbi) %>% arrange(desc(avg_hatecrimes_per_100k_fbi))

hate_crimes %>% select(state, share_voters_voted_trump, hate_crimes_per_100k_splc) %>% arrange(desc(hate_crimes_per_100k_splc))
```
Despite having the **lowest share of Trump Voters**, Washington D.C. has the **highest number of hate crimes**, both before the 2016 election and during election week 2016. Why might this be the case?

Let us examine what other variables may be significant in explaining why D.C. has the highest number of hate crimes.

```{r}
hate_crimes %>% select(state, median_household_income, hate_crimes_per_100k_splc) %>% arrange(desc(median_household_income))
```
Median Household Income: 5th

```{r}
hate_crimes %>% select(state, share_unemployed_seasonal, hate_crimes_per_100k_splc) %>% arrange(desc(share_unemployed_seasonal))
```
**Seasonal Unemployment: 3rd**

```{r}
hate_crimes %>% select(state, share_population_in_metro_areas, hate_crimes_per_100k_splc) %>% arrange(desc(share_population_in_metro_areas))
```

**Share living in Metro Areas: 1st**

```{r}
hate_crimes %>% select(state, share_population_with_high_school_degree, hate_crimes_per_100k_splc) %>% arrange(desc(share_population_with_high_school_degree))
```

Share of High School Graduates: 28th

```{r}
hate_crimes %>% select(state, share_non_citizen, hate_crimes_per_100k_splc) %>% arrange(desc(share_non_citizen))
```

**Share of Non-Citizens: 2nd**

```{r}
hate_crimes %>% select(state, share_white_poverty, hate_crimes_per_100k_splc) %>% arrange(desc(share_white_poverty))
```

**Share of White Poverty: 51st**

```{r}
hate_crimes %>% select(state, gini_index, hate_crimes_per_100k_splc) %>% arrange(desc(gini_index))
```

**Gini Index (higher is more unequal): 1st**

```{r}
hate_crimes %>% select(state, share_non_white, hate_crimes_per_100k_splc) %>% arrange(desc(share_non_white))
```

**Share of Non-Whites: 1st**

Based on the above, we hypothesise that **Seasonal Unemployment, Share living in Metro Areas, Share of Non-Citizens, Share of White Poverty, and Gini Index** may be **correlated** with number of Hate Crimes via _(a) variable(s) not captured in the dataset._


### Working Notes








